Rice yield improvements through plant breeding are offset by inherent yield declines over time

### 2018-02-21 output/breed_paper_revised.Rnw
Remove 'yield erosion'.

Meeting the challenge of feeding a growing population with limited
resources will require increasing the yield potential of staple crops,
such as rice. Yet many high-yielding, intensive production systems
have experienced slow rates of yield improvement in recent years
despite a demonstrated increase in the yield potential of new crop
cultivars. We analyzed experimental data from one such cropping
system, i.e., California (CA) rice, in order to quantify improvements
made in the genetic yield potential obtained through plant
breeding. California rice systems are among the highest in the world
and close to maximum yield potential. Specifically, the hypothesis was
tested that if rice cultivar yields decline over time (termed yield
erosion) then apparent yield increases in side-by-side yield
comparison tests will not reflect increases in yield potential. This
hypothesis was tested using 33 years of experimental yield data from
the California Cooperative Rice Research Foundation Rice Experiment
Station. Based on side-by-side comparisons of old and new rice
cultivars which do not consider yield erosion, there was an apparent
increase in yield. However, the yields of older cultivars eroded at an
estimated rate of 29.3 kg ha\tsups{-1} year\tsups{-1} (90\% credible
interval -4.4 to -53.3) after initial selection. Once this effect was
considered, the yield advantage of newer cultivars over old was
uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\% credible interval
-36.1 to 31.5). These results highlight 1) the importance of
continuous crop improvement and deployment of new cultivars simply to
maintain existing yields, and 2) to increase the genetic yield
potential, higher yield targets are needed. Importantly, when breeding
near the yield potential, despite the limited yield gains, significant
advances in improving quality and reducing crop duration have been
made.

@@

Certainly one question in looking at this situation is, ``do new
cultivars have increased yield potential''?  Like many breeding
programs, one of the goals of the CCRRF is to increase yields through
breeding cultivars with increased yield potential. However, when the
popular cultivar M-202 was released in 1985, it was reported to yield
an average of 11 Mg ha\tsups{-1}~\citep[14\% moisture content
(MC);][]{m202}. In 2015, the reported average yield of the newest
released cultivar M-209 was 10.8 Mg ha\tsups{-1}~\citep[14\%
MC;][]{m209}. However, in annual breeding reports and cultivar release
announcements, data based on yield performance in side-by-side cultivar
trials show that new cultivars usually out yield the older
cultivars. In other rice systems, a decrease in a cultivar's yield
performance over time has been observed~\citep{Peng1998, Peng2010,
Peng2000, Datta1995}, often termed yield erosion~\citep{Peng1998}. It
has been speculated that this yield erosion is due to the inability of
a cultivar to adapt to changing biotic and abiotic
conditions~\citep{Peng1998, Peng2010, Peng2000, Datta1995}. Yield
erosion may explain the apparent contradiction above, in which
side-by-side yield comparisons show yield improvements in cultivars
while overall there may be little change in yield potential. To
address this issue we tested the hypothesis that if rice cultivar
yields erode over time then apparent yield increases in side-by-side
yield comparison tests will not reflect the true magnitude of yield
changes over time.

@@

As part of the annual Statewide Variety Trials, newly developed
entries are tested against officially released cultivars (checks) at
the RES in several trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as these cultivars constitute approximately
90\% of the planted rice area in CA. Newly developed entries are
typically evaluated in these trials for 3 to 5 years prior to their
official release. For these cultivars planted in breeding trials at
the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations (replicated 3
to 4 times in a CRB design) were collected leading to a final data set
of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
representing \Sexpr{length(unique(mod_data[,'id']))} cultivars and
included every publicly released medium grain cultivar developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of public release and the time elapsed (in years) since release
was determined.

@@

To test the hypothesis, the influence of breeding and yield erosion
were quantified on yield over time using three nested Bayesian hierarchical
models. First, yield was modeled as a response to the number of years since
the cultivar was officially released and the year of release using a
mixed-effects linear regression. To account for similarities between
years and between cultivars, year and cultivar were included in the
model as random effects.

@@

Where $yield_i$ is the grain yield for plot $i$, the Intercept is the
average yield for the RES across cultivars and years, $Rate_{yield\
erosion\ i}$ is rate of yield erosion per ${years\ since\ release\ i}$
(negative for years prior to release, zero at the year of release),
$Rate_{yield\ improvement\ i}$ is the rate of yield improvement per
year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
year, and $\beta_{cultivar}$ is the random effect of cultivar. This
full model was modified to further explore yield dynamics over
time. For comparison to estimates where yield erosion is not taken
into account, the model was used as described above except omitting
the $Rate_{yield\ erosion\ i}$ term (Model 2). Lastly, a model was fit
to estimate if there were differences between cultivars in the rate of
yield erosion by adding a random slope by cultivar (Model 3).

@@

\subsection{Yield erosion}
\begin{figure}[H]

@@

plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Year since release",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(8.0, 12.5))
with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 ,
       b = yrs_coef/1000)
polygon(x = c(yrs_in, rev(yrs_in)),
        y = c(ints[1,],rev(ints[2,])),
        border = NA,
        col = rgb(0,0,0,0.1))
text(5, 8.0, bquote(.(yrs_coef)~kg~ha^-1~year^-1))
mtext("Adjusting for yield erosion", adj = 0)

@@

There was no evidence of consistent differences between cultivars in
the rate of yield erosion (Model 3), hence the model excluding varying
rates of erosion by cultivar (Model 2) is presented here. The yields
of each cultivar over time were evaluated relative to when the
cultivar was released using the full model which accounts for both
yield erosion and yield improvements over time. The average decrease
in yield over time (i.e., yield erosion) once a cultivar was selected
was estimated at \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} (90\%
CI: \Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig \ref{fig:RES_year};
Table \ref{tbl:coef}). The oldest cultivar in the trials
(\Sexpr{gsub("M","M-",unique(mod_data[which.min(mod_data[,"release_yr"]),"id"]))})
is estimated to have lost an average of
\Sexpr{round(yrs_coef*-36/1000,1)} Mg ha\tsups{-1} since its
release. Likewise, an estimated 560 kg ha\tsups{-1} yield
advantage of M-206 over M-202 (the current and former predominant
cultivars, respectively), can be attributed solely to the erosion
of M-202's yield over time (Figs \ref{fig:first_yrs} \&
\ref{fig:compare_yrs}). Further supporting this estimate of yield
erosion, both cultivars yielded similarly at their respective year of
release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
released 19 years prior to M-206 (Fig. \ref{fig:compare_yrs}; Table
\ref{tbl:ids}).

@@

@
\caption{The performance of two popular cultivars over time. The
  dashed line is the estimated yield erosion as estimated by the full
  model. Points ($\bullet$) represent the mean, and bars (|) the standard
  error of the observed data.}
\label{fig:compare_yrs}
\end{figure}

@@

When yield erosion is not accounted for in the analysis (Model 2),
there is estimated to be an increase per year of approximately
\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1} (90\% CI
\Sexpr{only_release_CI[1]} to \Sexpr{only_release_CI[2]})
(Fig. \ref{fig:wo_erosion}). However, the full model (Model 1) which
accounts for both the year of release and yield erosion indicates that
the estimated change in grain yield per year in the CCRRF's breeding
trials at the RES was \Sexpr{release_yr_coef} kg ha\tsups{-1}
year\tsups{-1}, with both small positive and negative trends within
the credible interval (90\% CI: \Sexpr{rel_CI[1]} to
\Sexpr{rel_CI[2]}) (Fig. \ref{fig:first_yrs}). According to model
comparison metrics (PS-LOO, WAIC), the full model (Model 1) is
estimated to have superior predictive accuracy compared to the model
which does not account for yield erosion (Model 2)

After correcting for yield erosion, some cultivars were higher
yielding compared to the overall average (as estimated via the random
intercept term in the model). The marginal estimate for the
performance of a cultivar averaged over the uncertainty in the other
parameters in the model (i.e., the random effect) shows three
cultivars with higher than average yields:
\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),colla=', ')}
(Figs. \ref{fig:first_yrs} \& \ref{fig:ids_comp}).

@@

with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
text(x = plot_data$release_yr-0.4, y = plot_data$yield_kg + 0.07,
     labels = gsub("M", "M-", row.names(plot_data)),
     pos = 4, cex = 0.75)
text(x = 2000, y = 8, bquote(.(round(fixef(RES_rel_fit)[2] * scl, 1))~kg~ha^-1~yr^-1))
mtext("Without adjusting for yield erosion", adj = 0)

@
\caption{The relationship between yield and the release year of medium
grain rice cultivars for the last 6 years
(\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated without
correcting for yield erosion over time.}
\label{fig:wo_erosion}
\end{figure}

@@

plot(mean ~ release_yr, data = plot_data2, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = "", #expression("Grain yield -corrected( Mg ha"^-1~")"),
     yaxt = "n",
     ylim = c(8, 12.5),
     xaxt = "n",
     xlim = c(1980,2020))
axis(1, at = seq(1980, 2020, by = 10))
abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
           (2000 * release_yr_coef)) / 1000,
       b = release_yr_coef/1000,
       lty = 2)
with(plot_data2, segments(x0 = release_yr, y0 = mean + se,
                         y1 = mean - se))
text(x = plot_data2$release_yr-0.4, y = plot_data2$mean + 0.07,
     labels = gsub("M", "M-", row.names(plot_data2)),
     pos = 4, cex = 0.75)
text(x = 2000, y = 8, bquote(.(release_yr_coef)~kg~ha^-1~yr^-1))
mtext("Adjusting for yield erosion", adj = 0)

@@

@
\caption{The relationship between yield and the release year of medium
  grain rice cultivars for the last 6 years
  (\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated by 
  correcting for yield erosion over time. Yield erosion was corrected
  for by adjusting each observed yield by estimated amount lost 
  (\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} after release).}
\label{fig:w_erosion}
\end{figure}
\end{comment}

@@

Not accounting for yield erosion, the estimated yield improvement of
\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:wo_erosion}) is lower than the 37 to 60 kg ha\tsups{-1}
year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
comparisons. However, when accounting for yield erosion and looking
across all cultivars, there is no conclusive evidence of increased
genetic yield potential in medium grain rice cultivars released by
CCRRF's breeding improvement program over the period
\Sexpr{min(mod_data[,'year'])} to \Sexpr{max(mod_data[,'year'])}. 
Grain yield over this period has stayed relatively constant at an average
\Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
ha\tsups{-1}. Since the estimated yield improvement spans from
\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}), both increases and
decreases in the genetic yield potential over the study period are
plausible. This analysis suggests that the CCRRF breeding program has
maintained but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures. This
conclusion holds true even when evaluating the cultivars which have
been breed for broad adaption and high yield potential (M-202, M-206
and M-105). Yields of these cultivars at time of release have all
yielded between
\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
and
\Sexpr{round(max(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
Mg ha\tsups{-1} with no indication of yields increasing
(Fig. \ref{fig:first_yrs}).

Yields of a particular cultivar may decline overtime for a number of
reasons. First, ~\citet{Nie2009} reported yield declines in rice due
to continuously flooded conditions over a long time period which could
be overcome with improved N management. This is likely not the case
here as the fields used in this study grow rice only one season a year
and tend to be non-flooded for the remainder of the year. Furthermore,
at this location and unlike most CA rice fields, the fields are left
fallow every other year. Second, changing climate may influence yields
with either warmer~\citep{van2013} or cooler~\citep{espe2017}
temperatures reducing yields. Again, this is unlikely to be the case
in this situation as ~\citet{espe2017} found no significant changes in
growing season temperatures (either high, low, or average temps) the
Sacramento Valley during the study period. Thirdly, small changes in
biotic and abiotic forces have previously been speculated as the cause
for yield erosion of the cultivar IR8 at IRRI~\citep{Peng1998,
Peng2000, Peng2010}. Since the cultivar's genetics remain stable from
the time of selection but the conditions the cultivar experiences over
time are continuously changing, accumulated changes too small to
reliably measure could plausibly lead to yield
erosion~\citep{Peng1998, Peng2000, Peng2010}. However, it is not clear
if this mechanism is responsible for the yield erosion of
\Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in this
analysis. Biotic stressors such as pests and diseases are low in this
Mediterranean climate. Furthermore, these fields are fallowed every
other year and the residues are burned, further lowering the potential
for pest and disease pressure.

It is striking that the estimated yield erosion is similar to that
reported by \citet{Peng2010} over a similar time span (15\% decline
over roughly 30 years). This suggests the mechanisms responsible could
potentially be endemic to all rice production systems. While efforts
should be made to identify the cause of erosion, this study highlights
both the need for maintenance of plant breeding and the difficulty of
increasing genetic yield ~\citep{espe2017} potential, especially in
systems such as CA where yield is approaching the physiological limit
\citep{espe16a}.

A goal of many breeding programs is to not only maintain the benchmark
yield of the previous ``standard'' but to surpass it, especially if we
are to meet the target of 50\% increased food production by 2050
\citep{Foley2011a}. Given the estimated yield erosion, to hit an
ambitious target of 500 kg ha\tsups{-1} (approximately 5\%) increase
in yield per release (with roughly 4 years between releases), a new
cultivar needs to yield an additional \Sexpr{round(yrs_coef*-4,-1)} kg
ha\tsups{-1} in side-by-side comparison in addition to the 500 kg
ha\tsups{-1} target improvement. Often side-by-side yield improvements
show an improvement in grain yield of 2 to 3\% per release, in which case
no absolute improvements in yield have been made beyond maintaining
yield.

@@

RES_coef = cbind(med_eff, paste(CIs[,1],CIs[,2], sep = " to "))
rownames(RES_coef) = c("Intercept (Mg ha\\tsups{-1})",
                       "Yield improvement (kg ha\\tsups{-1} year\\tsups{-1})",
                       "Yield erosion (kg ha\\tsups{-1} year\\tsups{-1})")
colnames(RES_coef) = c("Median", "90\\% Credible interval")

@@

@
\caption{The performance of two major cultivars over the period
\Sexpr{min(mod_data[,'year'])} -- \Sexpr{max(mod_data[,'year'])}. The
solid line is the estimated yield erosion (\Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1}), while the dashed line is the published
3-year average yield of M-202 at its release (1982-1984). Both
cultivars have a similar yield at their respective year of release,
and similar decreases over time.}
\label{fig:compare_yrs2}
\end{figure}
\newpage
\end{comment}

@@

dotchart(id_est[,1],
         labels = gsub("M", "M-", row.names(id_est)),
         pch = 16,
         xlim = c(min(id_est[,2]),max(id_est[,3])),
         xlab = expression("Estimated difference from average ( Mg ha"^-1~")"))
segments(x0 = id_est[,2], x1 = id_est[,3], y0 = 1:nrow(id_est))
abline(v = 0, lty = 2)
@
\caption{The marginal estimates for the difference between cultivars,
  averaged over the uncertainty in the fixed effect parameters (i.e.,
  yield erosion and annual yield gains).}
\label{fig:ids_comp}
\end{figure}

>>>>

Meeting the challenge of feeding a growing population with limited
resources will require increasing the yield potential of staple crops,
such as rice. Yet many high-yielding, intensive production systems
have experienced slow rates of yield improvement in recent years
despite a demonstrated increase in the yield potential of new crop
cultivars. We analyzed experimental data from one such cropping
system, i.e., California (CA) rice, in order to quantify improvements
made in the genetic yield potential obtained through plant
breeding. California rice systems are among the highest in the world
and close to maximum yield potential. Specifically, the hypothesis was
tested that if rice cultivar yields decline over time (termed yield
erosion) then apparent yield increases in side-by-side yield
comparison tests will not reflect increases in yield potential. This
hypothesis was tested using 33 years of experimental yield data from
the California Cooperative Rice Research Foundation Rice Experiment
Station. Based on side-by-side comparisons of old and new rice
cultivars which do not consider yield decline over time, there was an apparent
increase in yield. However, the yields of older cultivars eroded at an
estimated rate of 29.3 kg ha\tsups{-1} year\tsups{-1} (90\% credible
interval -4.4 to -53.3) after initial selection. Once this effect was
considered, the yield advantage of newer cultivars over old was
uncertain (-3.3 kg ha\tsups{-1} year\tsups{-1}, 90\% credible interval
-36.1 to 31.5). These results highlight 1) the importance of
continuous crop improvement and deployment of new cultivars simply to
maintain existing yields, and 2) to increase the genetic yield
potential, higher yield targets are needed. Importantly, when breeding
near the yield potential, despite the limited yield gains, significant
advances in improving quality and reducing crop duration have been
made.

@@

Certainly one question in looking at this situation is, ``do new
cultivars have increased yield potential''?  Like many breeding
programs, one of the goals of the CCRRF is to increase yields through
breeding cultivars with increased yield potential. However, when the
popular cultivar M-202 was released in 1985, it was reported to yield
an average of 11 Mg ha\tsups{-1}~\citep[14\% moisture content
(MC);][]{m202}. In 2015, the reported average yield of the newest
released cultivar M-209 was 10.8 Mg ha\tsups{-1}~\citep[14\%
MC;][]{m209}. However, in annual breeding reports and cultivar release
announcements, data based on yield performance in side-by-side cultivar
trials show that new cultivars usually out yield the older
cultivars. In other rice systems, a decrease in a cultivar's yield
performance over time has been observed~\citep{Peng1998, Peng2010,
Peng2000, Datta1995}, sometimes termed ``yield erosion''~\citep{Peng1998}. It
has been speculated that this yield decline over time is due to the inability of
a cultivar to adapt to changing biotic and abiotic
conditions~\citep{Peng1998, Peng2010, Peng2000, Datta1995}. Yield
erosion may explain the apparent contradiction above, in which
side-by-side yield comparisons show yield improvements in cultivars
while overall there may be little change in yield potential. To
address this issue we tested the hypothesis that if rice cultivar
yields erode over time then apparent yield increases in side-by-side
yield comparison tests will not reflect the true magnitude of yield
changes over time.

@@

As part of the annual Statewide Variety Trials, newly developed
entries are tested against officially released cultivars (checks) at
the RES in several trials spanning multiple planting dates per
year. For the purposes of this study, we concentrated on released
medium grain cultivars, as these cultivars constitute approximately
90\% of the planted rice area in CA. Newly developed entries are
typically evaluated in these trials for 3 to 5 years prior to their
official release. For these cultivars planted in breeding trials at
the RES from \Sexpr{min(mod_data[,'year'])} --
\Sexpr{max(mod_data[,'year'])} plot level observations (replicated 3
to 4 times in a RCB design) were collected leading to a final data set
of \Sexpr{format(nrow(mod_data), big.mark = ',')} observations
representing \Sexpr{length(unique(mod_data[,'id']))} cultivars and
included every publicly released medium grain cultivar developed by
the CCRRF from \Sexpr{min(mod_data[,'release_yr'])} --
\Sexpr{max(mod_data[,'release_yr'])}. For each cultivar, the official
year of public release and the time elapsed (in years) since release
was determined.

@@

To test the hypothesis, the influence of breeding and yield decline over time
were quantified on yield over time using three nested Bayesian hierarchical
models. First, yield was modeled as a response to the number of years since
the cultivar was officially released and the year of release using a
mixed-effects linear regression. To account for similarities between
years and between cultivars, year and cultivar were included in the
model as random effects.

@@

Where $yield_i$ is the grain yield for plot $i$, the Intercept is the
average yield for the RES across cultivars and years, $Rate_{yield\
erosion\ i}$ is rate of yield decline per ${years\ since\ release\ i}$
(negative for years prior to release, zero at the year of release),
$Rate_{yield\ improvement\ i}$ is the rate of yield improvement per
year (${release\ year\ i}$), $\beta_{year}$ is the random effect of
year, and $\beta_{cultivar}$ is the random effect of cultivar. This
full model was modified to further explore yield dynamics over
time. For comparison to estimates where yield decline over time is not taken
into account, the model was used as described above except omitting
the $Rate_{yield\ erosion\ i}$ term (Model 2). Lastly, a model was fit
to estimate if there were differences between cultivars in the rate of
yield decline over time by adding a random slope by cultivar (Model 3).

@@

\subsection{yield decline over time}
\begin{figure}[H]

@@

plot(yield_kg ~ yrs_in_trial, data = plot_data, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Year since release",
     ylab = expression("Grain yield ( kg ha"^-1~")"),
     ylim = c(8.0, 12.5))
with(plot_data, segments(x0 = yrs_in_trial, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
abline(a = ((fixef(RES_full)[1] * scl) + center) / 1000 ,
       b = yrs_coef/1000)
polygon(x = c(yrs_in, rev(yrs_in)),
        y = c(ints[1,],rev(ints[2,])),
        border = NA,
        col = rgb(0,0,0,0.1))
text(5, 8.0, bquote(.(yrs_coef)~kg~ha^-1~year^-1))
mtext("Adjusting for yield decline over time", adj = 0)

@@

There was no evidence of consistent differences between cultivars in
the rate of yield decline over time (Model 3), hence the model excluding varying
rates of erosion by cultivar (Model 2) is presented here. The yields
of each cultivar over time were evaluated relative to when the
cultivar was released using the full model which accounts for both
yield decline over time and yield improvements over time. The average decrease
in yield over time once a cultivar was selected
was estimated at \Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} (90\%
CI: \Sexpr{RES_CI[2]} to \Sexpr{RES_CI[1]}) (Fig \ref{fig:RES_year};
Table \ref{tbl:coef}). The oldest cultivar in the trials
(\Sexpr{gsub("M","M-",unique(mod_data[which.min(mod_data[,"release_yr"]),"id"]))})
is estimated to have lost an average of
\Sexpr{round(yrs_coef*-36/1000,1)} Mg ha\tsups{-1} since its
release. Likewise, an estimated 560 kg ha\tsups{-1} yield
advantage of M-206 over M-202 (the current and former predominant
cultivars, respectively), can be attributed solely to the erosion
of M-202's yield over time (Figs \ref{fig:first_yrs} \&
\ref{fig:compare_yrs}). Further supporting this estimate of yield
erosion, both cultivars yielded similarly at their respective year of
release (Fig. \ref{fig:compare_yrs}), despite the fact that M-202 was
released 19 years prior to M-206 (Fig. \ref{fig:compare_yrs}; Table
\ref{tbl:ids}).

@@

@
\caption{The performance of two popular cultivars over time. The
  dashed line is the estimated yield decline over time as estimated by the full
  model. Points ($\bullet$) represent the mean, and bars (|) the standard
  error of the observed data.}
\label{fig:compare_yrs}
\end{figure}

@@

When yield decline over time is not accounted for in the analysis (Model 2),
there is estimated to be an increase per year of approximately
\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1} (90\% CI
\Sexpr{only_release_CI[1]} to \Sexpr{only_release_CI[2]})
(Fig. \ref{fig:wo_erosion}). However, the full model (Model 1) which
accounts for both the year of release and yield decline over time indicates that
the estimated change in grain yield per year in the CCRRF's breeding
trials at the RES was \Sexpr{release_yr_coef} kg ha\tsups{-1}
year\tsups{-1}, with both small positive and negative trends within
the credible interval (90\% CI: \Sexpr{rel_CI[1]} to
\Sexpr{rel_CI[2]}) (Fig. \ref{fig:first_yrs}). According to model
comparison metrics (PS-LOO, WAIC), the full model (Model 1) is
estimated to have superior predictive accuracy compared to the model
which does not account for yield decline over time (Model 2)

After correcting for yield decline over time, some cultivars were higher
yielding compared to the overall average (as estimated via the random
intercept term in the model). The marginal estimate for the
performance of a cultivar averaged over the uncertainty in the other
parameters in the model (i.e., the random effect) shows three
cultivars with higher than average yields:
\Sexpr{paste(gsub('M','M-',rev(row.names(id_est))[c(1,2,4)]),colla=', ')}
(Figs. \ref{fig:first_yrs} \& \ref{fig:ids_comp}).

@@

with(plot_data, segments(x0 = release_yr, y0=yield_kg + yield_se,
                         y1 = yield_kg - yield_se))
text(x = plot_data$release_yr-0.4, y = plot_data$yield_kg + 0.07,
     labels = gsub("M", "M-", row.names(plot_data)),
     pos = 4, cex = 0.75)
text(x = 2000, y = 8, bquote(.(round(fixef(RES_rel_fit)[2] * scl, 1))~kg~ha^-1~yr^-1))
mtext("Without adjusting for yield decline over time", adj = 0)

@
\caption{The relationship between yield and the release year of medium
grain rice cultivars for the last 6 years
(\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated without
correcting for yield decline over time.}
\label{fig:wo_erosion}
\end{figure}

@@

plot(mean ~ release_yr, data = plot_data2, pch = 16, #col = rgb(0,0,0,0.25),
     xlab = "Release year",
     ylab = "", #expression("Grain yield -corrected( Mg ha"^-1~")"),
     yaxt = "n",
     ylim = c(8, 12.5),
     xaxt = "n",
     xlim = c(1980,2020))
axis(1, at = seq(1980, 2020, by = 10))
abline(a = ((((fixef(RES_full)[1] * scl) + center) ) -
           (2000 * release_yr_coef)) / 1000,
       b = release_yr_coef/1000,
       lty = 2)
with(plot_data2, segments(x0 = release_yr, y0 = mean + se,
                         y1 = mean - se))
text(x = plot_data2$release_yr-0.4, y = plot_data2$mean + 0.07,
     labels = gsub("M", "M-", row.names(plot_data2)),
     pos = 4, cex = 0.75)
text(x = 2000, y = 8, bquote(.(release_yr_coef)~kg~ha^-1~yr^-1))
mtext("Adjusting for yield decline over time", adj = 0)

@@

@
\caption{The relationship between yield and the release year of medium
  grain rice cultivars for the last 6 years
  (\Sexpr{min(comp_yrs)} -- \Sexpr{max(comp_yrs)}) as estimated by 
  correcting for yield decline over time. yield decline over time was corrected
  for by adjusting each observed yield by estimated amount lost 
  (\Sexpr{yrs_coef} kg ha\tsups{-1} year\tsups{-1} after release).}
\label{fig:w_erosion}
\end{figure}
\end{comment}

@@

Not accounting for yield decline over time, the estimated yield improvement of
\Sexpr{only_release_coef} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:wo_erosion}) is lower than the 37 to 60 kg ha\tsups{-1}
year\tsups{-1} estimated by~\citet{Omar2014} in side-by-side
comparisons. However, when accounting for yield decline over time and looking
across all cultivars, there is no conclusive evidence of increased
genetic yield potential in medium grain rice cultivars released by
CCRRF's breeding improvement program over the period
\Sexpr{min(mod_data[,'year'])} to \Sexpr{max(mod_data[,'year'])}. 
Grain yield over this period has stayed relatively constant at an average
\Sexpr{round(mean(mod_data[,"yield_kg"])/1000, 1)} Mg
ha\tsups{-1}. Since the estimated yield improvement spans from
\Sexpr{rel_CI[1]} to \Sexpr{rel_CI[2]} kg ha\tsups{-1} year\tsups{-1}
(Fig. \ref{fig:RES_year}; Table \ref{tbl:coef}), both increases and
decreases in the genetic yield potential over the study period are
plausible. This analysis suggests that the CCRRF breeding program has
maintained but not substantially improved the yield
potential in the face of changing biotic and abiotic pressures. This
conclusion holds true even when evaluating the cultivars which have
been breed for broad adaption and high yield potential (M-202, M-206
and M-105). Yields of these cultivars at time of release have all
yielded between
\Sexpr{tmp=mod_data[mod_data[,"yrs_in_trial"]<=0&mod_data[,"id"]%in%c("M202","M206","M105"),];round(min(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
and
\Sexpr{round(max(by(tmp,tmp[,"id"],function(x)mean(x[,"yield_kg"]))/1000),2)}
Mg ha\tsups{-1} with no indication of yields increasing
(Fig. \ref{fig:first_yrs}).

Yields of a particular cultivar may decline overtime for a number of
reasons. First, ~\citet{Nie2009} reported yield declines in rice due
to continuously flooded conditions over a long time period which could
be overcome with improved N management. This is likely not the case
here as the fields used in this study grow rice only one season a year
and tend to be non-flooded for the remainder of the year. Furthermore,
at this location and unlike most CA rice fields, the fields are left
fallow every other year. Second, changing climate may influence yields
with either warmer~\citep{van2013} or cooler~\citep{espe2017}
temperatures reducing yields. Again, this is unlikely to be the case
in this situation as ~\citet{espe2017} found no significant changes in
growing season temperatures (either high, low, or average temps) the
Sacramento Valley during the study period. Thirdly, small changes in
biotic and abiotic forces have previously been speculated as the cause
for yield decline over time of the cultivar IR8 at IRRI~\citep{Peng1998,
Peng2000, Peng2010}. Since the cultivar's genetics remain stable from
the time of selection but the conditions the cultivar experiences over
time are continuously changing, accumulated changes too small to
reliably measure could plausibly lead to yield
erosion~\citep{Peng1998, Peng2000, Peng2010}. However, it is not clear
if this mechanism is responsible for the yield decline over time of
\Sexpr{yrs_coef} Mg ha\tsups{-1} year\tsups{-1} found in this
analysis. Biotic stressors such as pests and diseases are low in this
Mediterranean climate. Furthermore, these fields are fallowed every
other year and the residues are burned, further lowering the potential
for pest and disease pressure.

It is striking that the estimated yield decline over time is similar to that
reported by \citet{Peng2010} over a similar time span (15\% decline
over roughly 30 years). This suggests the mechanisms responsible could
potentially be endemic to all rice production systems. While efforts
should be made to identify the cause of erosion, this study highlights
both the need for maintenance of plant breeding and the difficulty of
increasing genetic yield ~\citep{espe2017} potential, especially in
systems such as CA where yield is approaching the physiological limit
\citep{espe16a}.

A goal of many breeding programs is to not only maintain the benchmark
yield of the previous ``standard'' but to surpass it, especially if we
are to meet the target of 50\% increased food production by 2050
\citep{Foley2011a}. Given the estimated yield decline over time, to hit an
ambitious target of 500 kg ha\tsups{-1} (approximately 5\%) increase
in yield per release (with roughly 4 years between releases), a new
cultivar needs to yield an additional \Sexpr{round(yrs_coef*-4,-1)} kg
ha\tsups{-1} in side-by-side comparison in addition to the 500 kg
ha\tsups{-1} target improvement. Often side-by-side yield improvements
show an improvement in grain yield of 2 to 3\% per release, in which case
no absolute improvements in yield have been made beyond maintaining
yield.

@@

RES_coef = cbind(med_eff, paste(CIs[,1],CIs[,2], sep = " to "))
rownames(RES_coef) = c("Intercept (Mg ha\\tsups{-1})",
                       "Yield improvement (kg ha\\tsups{-1} year\\tsups{-1})",
                       "Yield decline (kg ha\\tsups{-1} year\\tsups{-1})")
colnames(RES_coef) = c("Median", "90\\% Credible interval")

@@

@
\caption{The performance of two major cultivars over the period
\Sexpr{min(mod_data[,'year'])} -- \Sexpr{max(mod_data[,'year'])}. The
solid line is the estimated yield decline over time (\Sexpr{yrs_coef} kg
ha\tsups{-1} year\tsups{-1}), while the dashed line is the published
3-year average yield of M-202 at its release (1982-1984). Both
cultivars have a similar yield at their respective year of release,
and similar decreases over time.}
\label{fig:compare_yrs2}
\end{figure}
\newpage
\end{comment}

@@

dotchart(id_est[,1],
         labels = gsub("M", "M-", row.names(id_est)),
         pch = 16,
         xlim = c(min(id_est[,2]),max(id_est[,3])),
         xlab = expression("Estimated difference from average ( Mg ha"^-1~")"))
segments(x0 = id_est[,2], x1 = id_est[,3], y0 = 1:nrow(id_est))
abline(v = 0, lty = 2)
@
\caption{The marginal estimates for the difference between cultivars,
  averaged over the uncertainty in the fixed effect parameters (i.e.,
  yield decline over time and annual yield gains).}
\label{fig:ids_comp}
\end{figure}
